---
title: "Extracting names of functions defined in a script with treesitter"
date: '2024-07-18'
slug: extract-function-names-treesitter
output: hugodown::hugo_document
---

Coming back from a conference, we might be excited to install and try out the cool things we have heard about.
I, going against the stream :fish:, decided to experiment with a tool I have *not* heard about last week, as I unfortunately missed [Davis Vaughan's talk about treesitter](https://www.youtube.com/watch?v=Gm0ikRBAfwc).
Nonetheless, I caught the idea that treesitter is a parser of code, R code in particular.
The treesitter _R package_ uses the tree-sitter _C library_.
There are awesome applications of treesitter among which :sparkles: code search for R on GitHub :sparkles: but I learnt to know a bit of treesitter by solving a boring use case.

## My use case: I won't copy-paste these function names

Have you noticed this nice error message you get if you try to use `dplyr::across()` directly?

```{r, error=TRUE}
dplyr::across()
```

Kirill Müller got the idea to offer a similar behaviour to some igraph functions that exist only for use only within the square operator.
Therefore I had to find all functions defined _within_ this function (whose body has been simplified here):

```
`[.igraph.vs` <- function(...) {
  # some code
  bla <- function(...) {
    # some code
  }
  blop <- function(...) {
    # some code
  }
}
```

On the script outline I get on the right of my script in RStudio IDE[^positron], I can see the function names (that would be `bla()` and `blop()` in the simplified chunk above) but not copy-paste from there.

I would know how to extract them with [xmlparsedata](/2024/05/15/refactoring-xml/) but I thought it might be an opportunity for me to have a look at treesitter.
I went through different emotions as a beginner, not all of them positive :sob:, but I did get the function names in the end! :muscle:

## Get the whole tree from the script

Below, I dutifully followed the example on the [package homepage](https://davisvaughan.github.io/r-tree-sitter/index.html#example):

- load the R language from the treesitter.r package, as far as I understand the only language available at this point for the R package.
- load the parser for that language.
- read in the text.
- parse the text.

A last step is getting the tree root node as node, because I will query the whole script, and you can only query _nodes_, not _trees_.

```{r}
language <- treesitter.r::language()
parser <- treesitter::parser(language)
text <- brio::read_lines("/home/maelle/Documents/rigraph/R/iterators.R") |>
  paste(collapse = "\n")
tree <- treesitter::parser_parse(parser, text)
node <- treesitter::tree_root_node(tree)
```

## Find the node of the parent function of interest

The documentation page of [`treesitter::query()`](https://davisvaughan.github.io/r-tree-sitter/reference/query.html) recommends reading the documentation of tree-sitter about the [query syntax](https://tree-sitter.github.io/tree-sitter/using-parsers#query-syntax).
That documentation features very useful examples.

Below I am looking for "binary_operator" whose left hand side is an identifier that I capture as "name", and whose right hand side is a "function_definition" I capture as "def".
For some reason if I did not capture "def" then I got less information about the node. :shrug:
I also use a predicate: the name of the function has to be equal to "`[.igraph.vs`".

To find out I was after a "function_definition", I parsed a few lines of made-up code to study the resulting tree.

```{r}
square_bracket_thing <- '
(
((binary_operator
  lhs: (identifier) @name
  rhs: (function_definition)) @def
  (#eq? @name "`[.igraph.vs`"))
  
)
'
square_bracket_query <- treesitter::query(language, square_bracket_thing)
```

Then I executed the query and extracted the node.
In reality this took a bit more trial and error.

```{r}
square_bracket_thing_captures <- treesitter::query_captures(square_bracket_query, node)
square_bracket_thing_captures
square_fn <- square_bracket_thing_captures$node[[1]]
```

At this point I was already very proud of my tiny win but I did not have the "children functions" yet!

## Find the functions defined within the parent

I first considered creating a complicated nested query but I found no example of that.
I did see someone telling StackOverflow they did a recursive query and for some reason that gave me the idea of parsing the text of the node, then look for functions in that text.

The query was simpler: looking for function definitions, only capturing the names on the left hand side.

```{r}
square_tree <- treesitter::parser_parse(parser, treesitter::node_text(square_fn))
square_node <- treesitter::tree_root_node(square_tree)
kiddos_source <- '
(
(binary_operator
  lhs: (identifier) @name
  rhs: (function_definition))
)
'
kiddos_query <- treesitter::query(language, kiddos_source)
square_bracket_thing_captures <- treesitter::query_captures(kiddos_query, square_node)
head(square_bracket_thing_captures)
```

After that I was able to get the _names_ of the children functions!
I was actually only interested in those whose names start with a dot as all the other ones are deprecated anyway.

```{r}
kiddos_functions <- purrr::map_chr(square_bracket_thing_captures$node, treesitter::node_text)
kiddos_functions[startsWith(kiddos_functions, ".")]
```

Tada! Now I can go fix the issue I was tasked with.

## Conclusion

In this post I report on my first encounter with the treesitter package for parsing R code.
Copy-pasting the few function names would surely have been faster, but sometimes you've got to sit and [learn something new](https://www.pipinghotdata.com/posts/2022-11-02-regular-intentional-and-time-boxed-yak-shaving/). :relaxed:


[^positron]: No, I have not installed Positron yet.