---
title: "Three (four?) R functions I enjoyed this week"
date: '2023-08-30'
slug: three-R-functions
output: hugodown::hugo_document
---

There are already three functions of note on a piece of paper on my desk, so it's time to blog about them!

## How does this package depend on this other package? [`pak::pkg_deps_explain()`](https://pak.r-lib.org/reference/pkg_deps_explain.html)

The [pak package by Gábor Csárdi](https://pak.r-lib.org) makes installing packages easier. 
If I need to start working on a package, I clone it, then run [`pak::pak()`](https://pak.r-lib.org/reference/pak.html#details) to install and update its dependencies.
It's a "convenience function" that is convenient for sure! 
Bye bye `remotes::install_deps()`.

Anyway, pak is truly a treasure trove, even for challenges that happen less often.
Earlier this week I was seeing an error in a GitHub Actions log due to a package that I didn't know was a dependency of the package I was working on.
It was clearly not a direct dependency, it was not listed in `DESCRIPTION`.
It was a dependency of a dependency and I couldn't guess which one it was...
Luckily one doesn't need to guess!
Although I didn't think of it right away, the [`pak::pkg_deps_explain()`](https://pak.r-lib.org/reference/pkg_deps_explain.html) is my friend in such cases[^1]!

For instance, if you are wondering why usethis depends on httr2, 

```{r}
pak::pkg_deps_explain("usethis", "httr2")
```

usethis depends on gh (client for GitHub API) that in turns depends on httr2.

Next time I encounter a similar problem I can only hope I'll remember about this function sooner!

## Where in the file are there non ASCII characters? `tools::showNonASCIIfile()`

If you get the dreaded R CMD check WARNING "Found the following file with non-ASCII characters" and can't see at once what character that is in the file, you don't need to comb through each line of code.
You can simply run `tools::showNonASCIIfile(<filename>)`.
After that an easy fix can be to replace then with the `\uxxxx` escape as indicated in the WARNING.
I'm not saying it's always that easy but that's what happened to me this week!
I found the correct escape by using a search engine.

## How do these two text files differ? `tools::Rdiff()` or `gert::git_diff_patch()`

I'm working on the [babeldown R package](https://docs.ropensci.org/babeldown/) that helps translate mostly Markdown files.
Support for first-time automatic translation is there, but a next step is to add support for automatic update of a translation based on a git commit.
Say I have a file in English and its translation in Spanish and a contributor edits part of the Spanish file.
We want to have a function update the English file based on this diff.
As part of that work I need to programmatically parse the difference between two text files.

I saw `gert::git_diff()` but I didn't find the patch to be easily parsable.

I had better luck with `tools::Rdiff()` especially after a GitHub search showed me an example using it with `Log = TRUE`.
Before that I was unsuccessfully trying `capture.output()` to capture what was printed... I should have read the arguments list more carefully.

```{r}
original_lines <- c(
  "# Title", "",
  "## Subtitle", "",
  "Some info", "",
  "First line of a paragraph",
  "Second line of a paragraph"
)

amended_lines <- c(
  "# BIG Title", "",
  "## Subtitle", "",
  "Some info", "",
  "First line of a paragraph",
  "Second line of a paragraph",
  "More info"
)

file1 <- withr::local_tempfile()
file2 <- withr::local_tempfile()

brio::write_lines(original_lines, file1)
brio::write_lines(amended_lines, file2)

diff <- tools::Rdiff(file1, file2, Log = TRUE)
diff
```

What I especially liked in this output, is the lines `<line-numbers-in-original-file><letter-indicated-status><line-numbers-in-new-file>` where the status can be a (added), d (deleted) or c (changed).

In comparison here's what gert gives me.

```{r}
dir <- withr::local_tempdir()
gert::git_init(dir)

brio::write_lines(original_lines, file.path(dir, "file.txt"))
gert::git_add("file.txt", repo = dir)
gert::git_commit("first commit", repo = dir)

brio::write_lines(amended_lines, file.path(dir, "file.txt"))
gert::git_add("file.txt", repo = dir)
commit_of_interest <- gert::git_commit("second commit", repo = dir)

cat(gert::git_diff_patch(commit_of_interest, repo = dir))

```

I'm not exactly sure yet whether `tools::Rdiff()` will be enough for my [use case](https://mastodon.social/deck/@maelle/110972627375018302) but it might!

## Conclusion

My sticky note mentioned `pak::pkg_deps_explain()`, `tools::showNonASCIIfile()` (what case is this :sweat_smile:) and `tools::Rdiff()`, that helped me make progress on R code earlier this week.
Time to break out a fresh sticky note and see what ends up on it!

[^1]: I first wrote it "pak_deps_explain" instead of "pkg_deps_explain". :grimacing: