---
title: "How to open files, folders, websites in R"
date: '2025-06-19'
slug: cover-modify-r-packages
output: hugodown::hugo_document
tags:
  - good practice
  - code style
  - useful functions
---

Coming to you from France, a post about [_Mise en place_](https://en.wikipedia.org/wiki/Mise_en_place) for R projects.
In a less francophone phrasing: to get to work on something you have to open that thing, be it a script or a project or a website.
The easier that is, the faster you get to work.
In this post I'll show a roundup of R functions and related tools for opening scripts, projects and websites for yourself or on behalf of the users of your code.

_Many thanks to [Hannah Frick](https://www.frick.ws/) for providing inspiration for some items of this post!_


## Open any file in the editor: `utils::file.edit()`, {cli}

If you write code that creates a file at `path` and then is supposed to open it for the user, no need for you to use, say, `rstudioapi::documentOpen(path)` that only works in [RStudio IDE](https://posit.co/products/open-source/rstudio/).
You can use a base R function, `utils::file.edit()`!
It will open the path in the default editor.
Without my setting up anything, in RStudio IDE that is RStudio IDE and in [Positron](https://drmowinckels.io/blog/2025/positron-debugging/) that is Positron.
From an R session in the terminal[^adventurous], that is the default editor of my system[^atom].

I was surprised to see that `usethis::edit_file()` uses an RStudio specific function when available:

```{r, eval=FALSE}
# https://github.com/r-lib/usethis/blob/4aa55e72ccca131df2d98fcd84fff66724d6250a/R/edit.R#L32C1-L36C4
if (rstudio_available() && rstudioapi::hasFun("navigateToFile")) {
  rstudioapi::navigateToFile(path)
} else {
  utils::file.edit(path)
}
```

According to the [commit that added this logic](https://github.com/r-lib/usethis/commit/9ab2582980f0f4a8a1d565dba00345ac7aa7e2a2) more than 7 years ago, "utils::file.edit opens in dialog" which I do not understand.
Maybe it's different depending on the OS? Maybe RStudio changed in the meantime?
Please tell me if you have any more information on this. :pray:

In any case I really enjoy `file.edit()` in codebases.
Interactively, I do not use is as often as, say, Positron's shortcut for navigating to files (Ctrl+P on my machine).

Last but not least, if you want to make it easy for the user to open a file, without opening it on their behalf, in messages emitted through the [cli package](https://blog.r-hub.io/2023/11/30/cliff-notes-about-cli/) you can use the [file class](https://cli.r-lib.org/reference/inline-markup.html#classes):

```{r, eval=FALSE}
cli::cli_alert_success("Hey go edit {.file config.toml} please!")
```
> _"If the terminal supports ANSI hyperlinks (e.g. RStudio, iTerm2, etc.), then cli creates a clickable link that opens the file in RStudio or with the default app for the file type."_

[^adventurous]: Look how adventurous I was when preparing this post! Running R from a terminal!
[^atom]: Which is... [Atom](https://en.wikipedia.org/wiki/Atom_(text_editor)), meaning I live in the past since its end-of-life occurred a few years ago. :ghost: I actually rarely use it.

## Open the script or test file of a script-test pair: `usethis::use_r()` and `usethis::use_test()`

Coming back to the wonderful usethis package, when you name your test files after your scripts as you should,
running [`usethis::use_test()`](https://r-pkgs.org/testing-basics.html#create-a-test) when focussed on an R script will open its test file, and conversely `usethis::use_r()` when focussed on a test file.
This is still how I switch between the two even if inside Positron.

## Open a project: `positron`, project launcher, {usethis}

This is all very good when within a project, but how do you enter your IDE in the first place?
How sad would it be to lose momentum before even launching a project?

When I used RStudio IDE I would navigate to the folder of interest then double-click on the `.Rproj` file.
Hannah Frick inspired me, when demoing Positron to me, to think a bit harder about how much I click. :sweat_smile:
I have adopted a more keyboard-based workflow since I switched to Positron, which I could have done with RStudio IDE.

One thing in particular that Hannah showed me that looks cool is the use of a [project launcher](https://positron.posit.co/rstudio-rproj-file.html#use-an-application-launcher).

Now, I myself still use the terminal. 
For instance if I clone a repository[^usethis], I'll then open it with the `positron` command.

Another way to open projects I use a lot is, on Positron as on Rstudio, clicking to the recent projects -- when the IDE is already up and running.
Unfortunately, I created and [currently](https://stateofther.netlify.app/post/saperlipopette/) [often](https://www.meetup.com/rbuenosaires/events/308338205/) use a [package that creates throwaway folders for practicing Git](https://docs.ropensci.org/saperlipopette/) which makes the list of recent projects utter rubbish. 🙃

Within saperlipopette itself, to create and open the exercice folder on behalf of the user I use `usethis::create_project()`, which handily opens it in Positron when run in Positron and RStudio when run in RStudio.

[^usethis]: I know I could use `usethis::create_from_github()` and sometimes do but not always... Maybe the message of this post is that my workflows are a big mess.

## Open an URL in the browser: `browseURL()` and {cli}

Sometimes you want the user of some code of yours to go admire or read a web page.

To open an URL in the default browser, you can use `utils::browseURL()`.
In all of the `usethis::browse_` functions, `utils::browseURL()` is what's used under the hood.
For instance:

```{r, eval=FALSE}
browseURL("https://masalmon.eu/post")
```

To make it easy to open an URL from a message, you can use the [URL class](https://cli.r-lib.org/reference/inline-markup.html#classes) of the cli package.
For instance:

```{r, eval=FALSE}
cli::cli_alert_info("Go read that blog! {.url https://masalmon.eu/post}!")
```

When getting that message, the user will simply have to click on the link to follow it (after trusting the domain in Positron).

## Open only in interactive sessions: `interactive()`, `rlang::is_interactive()`

If you use `utils::file.edit()` or `utils::browseURL()` in your code, you need to ensure the session is interactive.

Either

```{r, eval=FALSE}
if (interactive()) {
  file.edit("config.toml")
}
```

Or


```{r, eval=FALSE}
if (rlang::is_interactive()) {
  file.edit("config.toml")
}
```

Compared to `interactive()`, `rlang::is_interactive()` also checks whether knitr or testthat are in progress, and provides an escape hatch through the `rlang_interactive` option.

All of usethis functions that try to open or browse something on behalf of the user behave differently based on `rlang::is_interactive()`.
[`usethis::edit_file()`'s source](https://github.com/r-lib/usethis/blob/4aa55e72ccca131df2d98fcd84fff66724d6250a/R/edit.R#L20),
[`usethis::view_url()`'s source](https://github.com/r-lib/usethis/blob/4aa55e72ccca131df2d98fcd84fff66724d6250a/R/helpers.R#L107).

## Conclusion

In this post I summarized some tools for opening scripts, projects, URLs for yourself or on behalf of your user.
I'd like to add some words of conclusion beyond the summary table below,
but after so much talk of opening, I have trouble **closing** this post. :grin:

```{=html}
| Target to open          | Tool                                                                                              | Audience                                 |
| ----------------------- | ------------------------------------------------------------------------------------------------- | ---------------------------------------- |
| File                    | `utils::file.edit()`                                                                            | Users of your code                        |
| \-                      | IDE's shortcut                                                                                    | You -- get to know your IDE              |
| \-                      | [`.file` class of cli message](https://cli.r-lib.org/reference/inline-markup.html#classes) | Users of your code                        |
| R script from test file | `usethis::use_r()`                                                                              | You                                      |
| \-                      | IDE's shortcut                                                                                    | You -- get to know your IDE              |
| Test file from R script | `usethis::use_test()` | You                                      |
| \-                      | IDE's shortcut                                                                                    | You -- get to know your IDE              |
| Project (folder)        | Rproj file                                                                                        | RStudio IDE users who like clicking       |
| \-                      | List of recent projects within the IDE                                                            | You when the IDE is already launched     |
| \-                      | `rstudio`, `positron`                                                                         | Terminal dwellers                        |
| \-                      | [Project launcher](https://positron.posit.co/rstudio-rproj-file.html#use-an-application-launcher) | Positron users on macOS                  |
| \-                      | `usethis::create_project()`                                                                     | Users of your code that created a project |
| URL                     | `utils::browseURL()`                                                                            | Users of your code                        |
| \-                      | [`.url` class of cli message](https://cli.r-lib.org/reference/inline-markup.html#classes) | Users of your code                        |
```