---
title: "Useful functions for dealing with object names"
date: '2023-11-06'
slug: functions-dealing-with-names
output: hugodown::hugo_document
---

My sticky note filled up quickly after I only added `setNames()` on it, with related functions for dealing with object names, in base R and beyond!

## (Un)Setting object names: `stats::setNames()`, `unname()` and `rlang::set_names()`

I noticed a function ending with something like this:

```{r, eval=FALSE}
blop <- function() {
  
  # code creating the df data.frame
  # ...
  
  names(df) <- c("col1", "col2")
  df
}
```

It struck me as simplifiable by:

```{r, eval=FALSE}
blop <- function() {
  
  # code creating the df data.frame
  # ...
  
  stats::setNames(df, c("col1", "col2"))
}
```

Interestingly the docs for `setNames()` sound as if it were created just for this use case!

> "This is a convenience function that sets the names on an object and returns the object. It is most useful at the end of a function definition where one is creating the object to be returned and would prefer not to store it under a name just so the names can be assigned."

For the opposite operation, removing the names of an object, we can use `unname()`.

Unsurprisingly, the rlang package has its own special `rlang::set_names()`, that is more or less a `stats::setNames()` + `unname()` + generic `janitor::clean_names()` combo!

```{r}
(df <- data.frame(c(1, 2), c(5, 7)))

rlang::set_names(df, c("a", "b")) # sets names!

rlang::set_names(df, NULL) # removes names

df2 <- rlang::set_names(df, c("SHOUTING", "LOUDLY"))
df2
rlang::set_names(df2, tolower) # using a function to transform names

```

Of note, 

> "`set_names()` is stable and exported in purrr".

So if you prefer, you can use `purrr::set_names()` instead.

Last but not least, despite the examples above being on a data.frame, these functions can be used on lists, vectors...

## Extracting names: `names()`, `rlang::names2()`

`names()` probably doesn't need any introduction, but it's interesting to mention it in contrast to `rlang::names2()` that " always returns a character vector, even when an object does not have a names attribute.".

```{r}
vector <- c(1, 2)
names(vector)
rlang::names2(vector)
```

rlang also has a replacement variant, `rlang::names2<-`, that "never adds NA names and instead fills unnamed vectors with `""`."

I didn't know all this before skimming the list of rlang functions [dealing with object attributes](https://rlang.r-lib.org/reference/index.html#attributes).

## Checking names

### Is the object named: `rlang::is_named()`, `rlang::is_named2()`, `rlang::have_name()`


`rlang::have_name()` is the vectorized version of `rlang::is_name()`, not to be confused with `rlang::has_name()`. :sweat_smile:

```{r}
rlang::is_named(c(a = 1))

rlang::is_named(NULL)
rlang::is_named2(NULL)

rlang::have_name(c(a = 1, 2)) # the example from the manual page

```

I do not know of a base R equivalent, I mean I'd find a way to express it, but not as directly.
But maybe I am missing yet another base R gem!

### Does the object have an element called X, does the object have elements called X, Y: `utils::hasName()`, `rlang::has_name()`

If you want to check that an object contains elements with certain names, you can use `utils::hasName()`

```{r}
expected_names <- c("pof", "blop")

vector1 <- c(pof = 1, bla = 2)
vector2 <- c(pof = 1, bla = 2, blop = 3)

utils::hasName(vector1, expected_names)
utils::hasName(vector2, expected_names)
```


The rlang package has a similar function,  `rlang::has_name()`.

```{r}
expected_names <- c("pof", "blop")

vector1 <- c(pof = 1, bla = 2)
vector2 <- c(pof = 1, bla = 2, blop = 3)

rlang::has_name(vector1, expected_names)
rlang::has_name(vector2, expected_names)
```

I'm honestly not sure whether there's any difference between the utils and rlang versions, apart from the name!

### In package tests: `testthat::expect_named()`

`testthat::expect_named()` allows you to check that an object returned by your code has the names you expect, or simply has names.
You can ignore case and order based on the arguments `ignore.case` and `ignore.order`.

```{r, error=TRUE}
(object <- tibble::tibble(a = c(1, 2), b = c(3, 4)))

testthat::expect_named(object)

unnamed_object <- unname(object)
testthat::expect_named(unnamed_object)

testthat::expect_named(object, c("b", "a"))
testthat::expect_named(object, c("b", "a"), ignore.order = TRUE)

testthat::expect_named(object, c("A", "B"))
testthat::expect_named(object, c("A", "B"), ignore.case = TRUE)
```

If you're not used to using this expectation yet, lintr can [help you a bit](https://lintr.r-lib.org/reference/expect_named_linter.html).

## Conclusion

In this post I went through functions that deals with names: `stats::setNames()`, `unname()` and `rlang::set_names()` for (un)setting names (my initial motivation for this post!); `names()`, `rlang::names2()` for extracting names; `rlang::is_named()`, `rlang::is_named2()` and `rlang::have_name()` to find out whether an object is named; `utils::hasName()` and `rlang::has_name()` to find out whether an object contains elements of a given name; `testthat::expect_named()` for testing whether an object has names, or specific names.
Any other related function that comes to mind?