---
title: "Reducing my for loop usage with purrr::reduce()"
date: '2023-07-26'
tags:
  - good practice
  - code style
slug: reduce
output: hugodown::hugo_document
---

I (only! but luckily!) recently got introduced to the magic of `purrr::reduce()`. _Thank you, [Tobias](https://github.com/TSchiefer)!_
I was told about it right as I was unhappily using many for loops in a package[^glitter], for lack of a better idea. 
In this post I'll explain how `purrr::reduce()` helped me reduce my for loop usage. 
I also hope that if I'm doing something wrong, [someone will come forward and tell me](https://xkcd.com/386/)!

*This post was featured on the [R Weekly podcast](https://podverse.fm/episode/1odIoSlJl) by Eric Nantz and Mike Thomas.*

## Before: many for, much sadness

I was starting from a thing, that could be a list, or even a data.frame.
Then for a bunch of variables, I tweaked the thing.
My initial coding pattern was therefore:

```{r, eval=FALSE}
for (var in variables_vector) {
  thing <- do_something(thing, var, other_argument = other_argument)
}
```

I was iteratively changing the thing, along a `variables_vector`, or sometimes a `variables_list`.

### Silly example

Ugh, finding an example is hard, it feels very contrived but I promise my real-life adoption of `purrr::usage()` was life-changing!

```{r}
# Some basic movie information
movies <- tibble::tribble(
  ~title, ~color, ~elements,
  "Barbie", "pink", "shoes",
  "Oppenheimer", "red", "history"
)

# More information to add to movies
info_list <- list(
  list(title = "Barbie", info = list(element = "sparkles")),
  list(title = "Barbie", info = list(element = "feminism")),
  list(title = "Oppenheimer", info = list(element = "fire"))
)

# Don't tell me this is weirdly formatted data,
# who never obtains weirdly formatted data?!
info_list

add_element <- function(movies, info) {
  movies[movies[["title"]] == info[["title"]],][["elements"]] <-
    toString(c(
      movies[movies[["title"]] == info[["title"]],][["elements"]],
      info[["info"]][[1]]
    ))
  movies
}
```

Now how do I add each element of the list to the original table?
I could type something like:

```{r}
for (info in info_list) {
  movies <- add_element(movies, info)
}
movies
```

It's not too bad, really.
But since there's another way, we can change it.

## After

With `purrr::reduce()`

```{r, eval=FALSE}
for (var in variables_vector) {
  thing <- do_something(thing, var)
}
```

can become

```{r, eval=FALSE}
thing <- purrr::reduce(variables_vector, do_something, .init = thing)
``` 

And (notice the other argument),

```{r, eval=FALSE}
for (var in variables_vector) {
  thing <- do_something(thing, var, other_argument = other_argument)
}
```

can become

```{r, eval=FALSE}
thing <- purrr::reduce(
  variables_vector, 
  \(thing, x) do_something(thing, x, other_argument = other_argument), 
  .init = thing
)
``` 

I haven't completely internalized the pattern above but the [documentation of `purrr::reduce()`](https://purrr.tidyverse.org/reference/reduce.html#arguments) states

> "We now generally recommend against using ... to pass additional (constant) arguments to .f. Instead use a shorthand anonymous function:
> 
> Instead of
> x |> map(f, 1, 2, collapse = ",")
> do:
> x |> map(\(x) f(x, 1, 2, collapse = ","))
> This makes it easier to understand which arguments belong to which function and will tend to yield better error messages."

It might remind you of how things work for `dplyr::across()` [these days](https://mastodon.social/deck/@blasbenito@fosstodon.org/110745684166845628).

### Back to our silly example!

```{r}
# Some basic movie information
movies <- tibble::tribble(
  ~title, ~color, ~elements,
  "Barbie", "pink", "shoes",
  "Oppenheimer", "red", "history"
)

# More information to add to movies
info_list <- list(
  list(title = "Barbie", info = list(element = "sparkles")),
  list(title = "Barbie", info = list(element = "feminism")),
  list(title = "Oppenheimer", info = list(element = "fire"))
)

add_element <- function(movies, info) {
  movies[movies[["title"]] == info[["title"]],][["elements"]] <-
    toString(c(
      movies[movies[["title"]] == info[["title"]],][["elements"]],
      info[["info"]][[1]]
    ))
  movies
}

purrr::reduce(info_list, add_element, .init = movies)
```

If we tweak the `add_element()` function to add a `separator` argument to it,

```{r}
add_element <- function(movies, info, separator) {
  movies[movies[["title"]] == info[["title"]],][["elements"]] <-
    paste(c(
      movies[movies[["title"]] == info[["title"]],][["elements"]],
      info[["info"]][[1]]
    ), collapse = separator)
  movies
}

purrr::reduce(
  info_list, 
  \(movies, x) add_element(movies, x, separator = " - "), 
  .init = movies
)

purrr::reduce(
  info_list, 
  \(movies, x) add_element(movies, x, separator = " PLUS "), 
  .init = movies
)
```

And voilà!

## Conclusion

In this post I presented my approximate understanding of `purrr::reduce()`,
that helped me avoid writing some for loops and instead more elegant code... or at least helped me understand a pattern that in the future I could use elegantly.
I can only hope I `purrr::accumulate()` more experience, as I very much still feel like a newbie.

For more information I'd recommend reading the documentation of `purrr::reduce()` to be aware of other features, [the content on the reduce family in Advanced R by Hadley Wickham](https://adv-r.hadley.nz/functionals.html#reduce)... and release-watching the purrr repo to keep up-to-date with latest recommendations.
You can also use GitHub Advanced Search to find examples of usage of the function in, say, [CRAN packages](https://github.com/search?q=purrr%3A%3Areduce+org%3Acran&type=code).

Edit: For another take of / use case of `purrr::reduce()`, June Choe wrote a nice detailed tutorial ["Collapse repetitive piping with reduce()"](https://yjunechoe.github.io/posts/2020-12-13-collapse-repetitive-piping-with-reduce/).

[^glitter]: The package is [glitter](https://github.com/lvaudor/glitter), where we store query objects as a list.