---
title: 'Storrrify #satRdayCDF 2018'
date: '2018-06-26'
tags:
  - rtweet
  - conference
  - satRday
  - Twitter
slug: storrrify-satrdaycdf-2018
---

On Saturday I was at my second [satRday](http://satrdays.org/) conference this year, lucky me! I got to attend [satRday Cardiff](http://cardiff2018.satrdays.org/) which was a great experience. I gave a talk about rOpenSci onboarding system of packages, find my slidedeck [here](https://maelle.github.io/satrday_cardiff/slides#1) and other slidedecks [at this address](https://github.com/satRdays/cardiff2018_talks). A lot of R goodness!

[As I did in March for satRday Cape Town](/2018/03/24/satrdaycapetown/), I'll use my own tweets to summarize the event, but this time, having switched my website to [blogdown](https://bookdown.org/yihui/blogdown/) I can use [Hugo shortcodes](https://gohugo.io/content-management/shortcodes/) as [recommended by Romain François](https://twitter.com/romain_francois/status/977535090308866048)!

<!--more-->

# A few words about the conference before storrifying away

One of my [part-time employers](/bio/), [Locke Data](https://itsalocke.com/), was a big sponsor of the event, so it had a nice stand including a giant chibi Steph Locke! I therefore got to meet my co-workers in the flesh, and several other people I had only interacted with online. I really enjoy socializing with fellow R addicts once in a while! The organisers did a great job, and made sure everyone felt welcomed. Maybe they had had a look at [Steph Locke's list of awesome event organiser resources ](https://github.com/stephlocke/awesome-organiser-resources)! In any case, thanks [David Parr](https://github.com/daveparr) and [Paul Brennan](https://github.com/brennanpincardiff)!

Now, let's have a look at the day as tweeted by me...

# Getting my tweets

This part using [`rtweet`](http://rtweet.info/) is similar to what I had done in March, except the hashtag changed.

```{r, cache = TRUE}
my_tweets <- rtweet::get_timeline(user = "ma_salmon", n = 5000)
my_satrday_tweets <- dplyr::filter(my_tweets,
                                   stringr::str_detect(text, "[sS]at[rR]day[cC][dD][fF]"),
                                   !is_retweet)
my_satrday_tweets <- dplyr::arrange(my_satrday_tweets, created_at)
```

I obtained `r nrow(my_satrday_tweets)` tweets.

# Embedding the tweets

As stated in the intro I used Hugo shortcodes: cf [this section of the `blogdown` book](https://bookdown.org/yihui/blogdown/content.html#shortcode). I thought it'd be easy but having several tweets made it hard somehow: even when playing with chunk options and `purrr::map`-ping and friends, I wasn't able to use `blogdown::shortcode` so I used its insides instead. Still, using Hugo shortcodes is easier than hacking html. _Edited to add: after I [opened an issue with the feature request](https://github.com/rstudio/blogdown/issues/307) in `blogdown` dev version there's now a `shortcodes` function, vectorized version of `shortcode`! Thanks Yihui! I changed the code below._

Enjoy reading, and if you can, attend one of the next satRday conferences, for instance in [Amsterdam on September the 1st](http://amsterdam2018.satrdays.org/)! Or [run one yourself](https://twitter.com/satRdays_org/status/1011627085507506177)?


```{r}
blogdown::shortcodes('tweet', my_satrday_tweets$status_id)
```