---
title: "Stingy Beanie baby webscraping"
date: '2021-01-15'
tags:
  - polite
slug: beanie-baby
output: hugodown::hugo_document
---

I've just finished teaching [blogging with R Markdown at R-Ladies Bangalore](/talks/2021-01-15-blogging-r-markdown/).
This has two consequences: I need to calm down a bit after the stress of live demoing & co, and I am inspired to, well, blog with R Markdown!
As I've just read a [fascinating book about the beanie baby bubble](https://www.goodreads.com/book/show/20821185-the-great-beanie-baby-bubble) and as I've seen rvest is getting an update, I've decided to harvest [Beaniepedia](https://beaniepedia.com).
Both of these things show I spend too much time on Twitter, as the book has been tweeted about [by Vicky Boykis](https://twitter.com/vboykis/status/1347575684802240512), and the package changes have been tweeted about [by Hadley Wickham](https://twitter.com/hadleywickham/status/1347260116932976643).
I call that [staying up-to-date](/2019/01/25/uptodate/), of course.

So, as a little challenge for today, what are the most common animals among Beanie babies?
Do I even need much webscraping to find this out?

## How does one webscrape these days

I've always liked webscraping, as I think I enjoy getting and transforming data more than I enjoy analyzing it.
Compared to my former self, 

* I know a bit more about XPath (starting with knowing it exists!) so I don't use regular expression to parse HTML.
* I use [polite](https://dmi3kno.github.io/polite/) for polite webscraping!
* I know it's best to spend more time pondering about strategies before hammering requests at a website.

As to rvest recent changes, I had a quick look at the [changelog](https://rvest.tidyverse.org/news/index.html) but since I hadn't used it in so long, it's not as if I had any habit to change!

## How to harvest Beaniepedia

I've noticed Beaniepedia has a [sitemap for all beanies](view-source:https://beaniepedia.com/beanies/sitemap.xml) so from that I can extract the URLs to all Beanie pages. 
That's a necessary step.

Now from there I could either

* Scrape each of this page, respectfully slowly, and extract the table that includes the beanie's information;
* Use a more frugal strategy by parsing URLs. E.g. from the path of `https://beaniepedia.com/beanies/beanie-babies/january-the-birthday-bear-series-2/` I can extract the category of the Beanie (a beanie baby as opposed to, say, an attic treasure) and the animal by splitting `january-the-birthday-bear-series-2` into pieces and see whether one is an animal. How would I recognize animals? By extracting the word coming after "the".

I'll choose the second strategy and leave the first one as an exercise to the reader. :wink:

## From XML to animal frequencies

Let's get to work! 
A _sine qua non_ condition is obviously the website being ok with our scraping stuff. 
The polite package would tell us whether the robots.txt file were against our doing this, and I also took time looking whether the website had any warning.
I didn't find any so I think we're good to go.

```{r}
session <- polite::bow(
  url = "https://beaniepedia.com/beanies/sitemap.xml",
  user_agent = "Maëlle Salmon https://masalmon.eu"
  )
sitemap <- polite::scrape(session)
sitemap
```

The `sitemap` object is an XML document. 
I will extract URLs with the xml2 package.

```{r}
sitemap <- xml2::xml_ns_strip(sitemap)
urls <- xml2::xml_text(xml2::xml_find_all(sitemap, ".//loc"))
head(urls)
```

Now I need to parse the URLs.
In an URL path like `beanies/beanie-babies/jerry-the-minion-2/` the second part is the category, the third part is the Beanie Baby name. 
I, as if I were a good collector :upside_down_face:, am not interested in Attic treasures, only in Beanie Babies.

```{r}
urls_df <- urltools::url_parse(urls)
urls_df <- dplyr::filter(urls_df, stringr::str_detect(path, "beanie-babies"))
```

This gives me `r nrow(urls_df)` Beanie babies. Let's parse the last part of their path.
An earlier attempt ignored that some Beanie babies don't have any "the" in their names, e.g. the Hello Kitty ones.
This is a limitation of my stingy approach.
The error messages by dplyr were most helpful! "The error occurred in row 185." is so handy!

```{r}

get_animal <- function(parsed_path) {
  
  if (all(unlist(parsed_path) != "the")) {
    return(NA)
  }
  
  animals <- unlist(parsed_path)[which(unlist(parsed_path) == "the") + 1]
  animals[length(animals)] # thanks, "The End the bear"
}

library("magrittr")
animals_df <- urls_df %>%
  dplyr::rowwise() %>%
  dplyr::mutate(parsed_path = stringr::str_split(path, "/", simplify = TRUE)[1,3]) %>%
  dplyr::mutate(parsed_path = stringr::str_split(parsed_path, "-")) %>%
  dplyr::mutate(animal = get_animal(parsed_path))
```

Now we're getting somewhere!


```{r}
dplyr::count(
  animals_df,
  animal,
  sort = TRUE
)
```

Is this result surprising? Probably not!
Now, let's have a look at the ones we did not identify.

```{r}
animals_df %>%
  dplyr::filter(is.na(animal)) %>%
  dplyr::pull(path)
```

Fair enough, and nothing endangering our conclusion that bears win.

## Conclusion

In this post I set out to find out what animals are the most common among Beanie babies.
I thought I'd freshen my rvest-ing skill but thanks to the sitemap, that's my rusty dplyr knowledge I was able to update a bit.
In the end, I learnt that `r round(sum(animals_df$animal=="bear", na.rm=TRUE)/nrow(animals_df), digits=2)*100`% of Beanie babies, at least the one registered on Beaniepedia, are bears.
Thanks to Beaniepedia maintainer for allowing this fun!
